Global tradeoffs between prosperity, inequality, and the environment

### 2022-06-23 paper/manuscript.Rmd
incorporating authors comments

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
# author: |
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
abstract: |
  Sustainability is often framed as achieving prosperity, equality, and low environmental impact. But what if there is a trilemma that inhibits the joint achievement of these three key goals? Here we analyzed the synergies and tradeoffs between these goals by investigating historical data on per-capita gross national income (prosperity), the Gini coefficient for income distribution (inequality), and per-capita ecological footprint (environmental impact). We find that no country in the world have successfully solved the trilemma, but there are common typologies that can inform common barriers and strategies to solve it. Some countries are trapped on a high inequality space, while prosperous countries can reduce inequality but typically at expenses of high environmental impacts. Achieving ambitous sustainable goals will require redistribution of wealth and responsibility for the stewardship of the planet.

@@

Sustainability has often been characterised as having three essential, nested dimensions: economy, society, and the environment. More specifically, sustainability has been defined as attaining high material living standards (prosperity), that are also well distributed (equality) and maintainable over the long run (achieved through environmental integrity”) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Although each is undeniably desirable for development, what if all three goals are not simultaneously achievable? It has often been assumed that the ends to which individuals as well as societies aspire inevitably cohere, but *a priori*, there is no reason to believe that diverse goals must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007;
@Greenspan:2001].

In the literature on development, such macro-level incompatibilities are not novel. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination
[@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011]. Additionally, researchers suggest that based on international trends, there may be persisting tradeoffs between changes in improving diet, mitigating diseases, and environmental integrity [@Clark:2018]. Another such “trilemma” that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.

For most of human history, scarcity has been a ubiquitous condition of life: material want, not sufficiency, was the lot of most people [@Maddison:2007;
@Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, approximately 10% of the world’s population still lives under the international poverty line (less than US\$ 1.90 dollars per day), with a much larger share subsisting above that threshold but still aspiring for higher and more secure living standards
[@TheWor:2018]. In a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world.

At the same time, while many average incomes for countries in what was once called the “Third World” have progressively converged with those of advanced, mostly Western, economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
The widespread feelings of injustice that such unequal distribution arouses has historically been called the "Social Question"[@Faist:2009]. Over the past decade, inequality has become an ever-urgent issue for policymakers as well as the general public [@Chancel:2020]. This urgency has been strengthened by a growing body of research demonstrating links between inequality and human wellbeing, health, environmental quality, and economic complexity [@Hamann:2018; @Leach:2018;
@Pikett:2014; @Wilkinson:2009; @Hartmann:2017].

Environmental conditions form the third, and arguably most fundamental, dimension of the sustainability trilemma. There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere [@Díaz:2019]. Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
[@Díaz:2019]. At the same time, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017]. Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].

Although all three goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two. Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly prioritize prosperity, equality, and the environment [@Sachs:2015]. In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by a given country and by the world as a whole. In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, and if it does exist, where does it occur and why does it occur there?

@@

To answer these questions, we examined the historical performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above  (Table \ref{tab1}). We chose the country scale due to practical considerations
of data availability, but also on the recognition that many of the most
consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018]. We operationalized the trilemma as the movement of countries overtime in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g. reducing inequality and increasing prosperity) There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies are movements along
the (blue) diagonal where the achievement of two goals can be positively correlated;
while tradeoffs are represented by the (red) axis where the achievement of one
pervades the achievement of another (negative correlations). It is important to note that in this framing, the sign of the correlation changes depending on the term used: i.e., if one speaks of equity as opposed to inequality, or environmental quality as opposed to environmental footprint. In the remainder of this paper we use inequality and environmental footprint because that is how the original data was structured. Therefore, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 4in, height = 4in]{conceptual_fig}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality while increasing prosperity and environmental quality. There could be synergies and trade-offs in achieving these three objectives. We explore trajectory typologies on the trilemma space.}
\label{fig:trilemma}
\end{figure*}
```

To quantify prosperity, equality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used
per-capita gross national income (GNI) in US\$ values.GNI is a cognate measure of gross domestic product (GDP) and the two are close in value for most countries. A key difference is that GNI accounts for earnings from foreign investments. The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source:
<http://data.worldbank.org>). The threshold for being "prosperous" was   US\$12,746 per capita per year, which is the World Bank’s definition for a high-income country [@TheWor:2018].

To measure inequality, we used the Gini coefficient for income. The Gini coefficient is a statistical measure of dispersion and has a long history of being used to measure the level of economic inequality within a country [@Pyatt:1976]. Hypothetically, a country with a Gini value of 0 is absolutely equal, while a value of 1 represents absolute inequality. Gini coefficients of income were taken from the World Income Inequality Database v. 3.4 (source: [https://www.wider.unu.edu/data](https://www.wider.unu.edu/data)). We used 0.30 as the threshold below which countries are identified as having low inequality, similar to findings from a historical survey by Piketty [@Piketty:2014].

We used per-capita ecological footprint to measure environmental impact. Ecological footprint is a holistic estimate of the biophysical resources used by human activity, in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998]. Its unit is the "global hectare", combining measurements of carbon emissions, built-up area, fishing grounds, croplands, forests, and grazing land into an index of environmental impact. In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological “deficit”. Though it is not exhaustive, the ecological footprint provides the most widely-used and data-rich approximation of overall human impact on the environment, and it accounts for the effects of trade on resource consumption by country. Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)). We used 1.68 global per capita hectares as a threshold for environmental impact, as it reflects the amount of average land needed to satisfy human needs without exceeding the global biocapacity to recover. In other words, it is the limit at which current generations’ needs can be satisfied without compromising the ability of future generations to do so. Ecological footprint estimates above 1.68 global per capita hectares, therefore, represent unsustainable levels of environmental impact.

@@

We used these data to chart the movement of countries across dimensions of prosperity, equality, and environmental integrity over time. We further investigated whether these movements trace out typologies of countries in the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters to fit. We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization purposes, and vector fitting to show in the principal component space how different features of the trilemma space pull clusters apart. All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).

An analysis of the missing values of the publicly available data shows that a assessing the change over time in the trilemma spaces is only possible for 53 countries. The most incomplete dataset is the inequality data (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemmma space (Fig \ref{fig:cube_time}).

@@

(Aim 1200w | 610w)

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{cube_time}
\caption{\textbf{Countries trajectories in the trilemma space} 53 trajectories from 1995 to 2017 are depicted for countries with $\leqslant 30\%$ missing values. The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (A), inequality and prosperity (B), and ecological footprint and prosperity (C). Thresholds that delimit the objective space are depicted as dotted lines.}
\label{fig:cube_time}
\end{figure*}
```

No country in the data simultaneously achieved high prosperity, low inequality, and high environmental standards. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between inequality and ecological footprint (Fig \ref{fig:cube_time}A - as the Gini indicator for income decreases, ecological footprint increases) and ecological footprint and prosperity (Fig \ref{fig:cube_time}C- ecological footprint moves in the same direction as per-capita GNI). Additionally, prosperity and inequality (Fig \ref{fig:cube_time}B) were also correlated, with a number of high-income countries (e.g., Sweden, New Zealand) having low levels of inequality in the past, but then showed high variation and a trend toward higher inequality levels in more recent years.

While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typology. To further investigate individual trajectories, we calculated the displacement on the trilemma space for each pair-wise dimension, where $\gamma$ is the angle in the inequality and prosperity plane while $\theta$ is the angle in the inequality and ecological footprint plane. These two angles and the total distance covered in the three-dimensional space are used to uncover country typologies.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{trilema_ensamble}
\caption{\textbf{Trilemma attractors} }
\label{fig:cube_ensamble}
\end{figure*}
```

There are certain regions of the trilemma space where countries tend to dwell (Fig \ref{fig:cube_ensamble}). The inequality and ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity and ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality and prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three clusters of countries.  Despite this clustering, countries showed distinctive profiles in terms of their direction of movement in the trilemma space (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's inequality and ecological footprint worsened while GNI per capita grew significantly. And while in Ukraine inequality and environmental impact actually fell, prosperity only increased marginally. To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories of countries in the trilemma space from 1995 to 2017 converged into three general types (Fig \ref{fig:pca}). First, there was a group of countries trapped in a region of high inequality, mainly in Latin America. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$, that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). The second group showed large displacements within the trilemma space but with no definite direction, characterized by particularly high variations in Gini and per-capita GNI (Fig \ref{fig:pca}A, in yellow). The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). If we ignore temporal dynamics and focus on the mean and standard deviation as summary statistics, we are able to include more countries (N=140). Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between inequality, prosperity, and the environment. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies, while the second (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{pca_trilemma}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. 3 clusters are identified when looking at the historical trajectory of countries (N=53) from 1995-2017 (A). 2 clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness. It complements the dynamic analysis with one summary statistic over time for 140 countries (B). }
\label{fig:pca}
\end{figure*}
```

@@

(Aim 1000 | 1472)

Our analyses found that inequality and ecological footprint goals pull countries in opposite directions (Fig \ref{fig:pca}) providing strong evidence for the difficulty of achieving both goals at the same time. This suggests the existence of regions within the trilemma space that are hard to escape, characterized by high levels of inequality. The historical data shows that countries typically reduce inequality and then increase ecological footprint, not the other way around (Fig \ref{fig:cube_ensamble}A). Importantly, many middle-income countries have similar levels of inequality as high-income ones, but at lower levels of ecological footprint and per-capita GNI (Fig \ref{fig:cube_ensamble}B, C). This means that countries can achieve low levels of inequality while attaining and preserving high environmental quality. Historical evidence shows that higher prosperity often leads to higher ecological footprints. Nonetheless, it is important to note that past trajectories do not necessarily determine plausible future pathways in the trilemma space.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 1.5in]{maps_trilemma}
\caption{\textbf{Countries mean performance on trilemma space} Mean values of ecological footprint (A), inequality (B), and gross national income (C). }
\label{fig:map}
\end{figure*}
```
The potential mechanisms of the tradeoffs and synergies between economic, social, and environmental goals remain open questions. Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the later grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. Since this relationship was first posited in 1955, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020]. This has made the identification of mechanisms difficult.  Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through systemic regularities of the global economy or proactive policies by national governments [@Pikett:2014]. In our analysis of the trilemma space, we did not find evidence for a KC-like relationship.

In terms of the economy-environment interface, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship and implicit explanation: countries increase their environmental footprints as incomes rise (which in modern history has been driven by industrializatoin and urbanization); but after reaching an inflection point of affluence, the footprints of an increasingly rich society decline (Barbier 1997). As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not (Dasgupta et al. 2002; Stern 2004). In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC. Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns (Mäler 2013). But what is clear from the wealth of diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.

>>>>

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
# author: |
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
abstract: |
  Sustainability is often framed as achieving prosperity, equality, and environmental integrity: in other words, high living standards, evenly-distributed, and low-impact. But what if there is a trilemma that inhibits the joint achievement of these three key goals? We analyzed synergies and tradeoffs between these dimensions of sustainability by investigating historical data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint. We found that no country in the world successfully solved this trilemma, but there are typologies of countries that can help identify common barriers, and strategies to overcome them. Countries typically reduce inequality and then increase prosperity and environmental impact. Some countries are trapped in a high inequality space, while prosperous countries have reduced inequality only at the expense of increasing environmental impact. Escaping this trilemma to achieve ambitious sustainability goals will likely require economic redistribution and more urgent collective action for planetary stewardship.

@@

Sustainability has often been characterised as having three essential, nested dimensions: economy, society, and the environment. More specifically, sustainability has been defined as attaining high material living standards (prosperity) that are well distributed (equality) and maintainable over the long run (achieved through environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Although each is a desirable goal of development, what if all three are not simultaneously achievable? It has often been assumed that the positive ethical ends to which individuals as well as societies aspire can cohere, but *a priori*, there is no reason to believe that diverse goals, however separately desirable, must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007;
@Greenspan:2001].

In the literature on development, such macro-level incompatibilities are not novel. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination
[@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011]. Additionally, researchers suggest that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018]. In this light, we argue that another such "trilemma" that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.

For most of human history, scarcity has been a ubiquitous condition of life: material want, not sufficiency, was the lot of most people [@Maddison:2007;
@Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, approximately 10% of the world’s population still lives under the international poverty line of US\$ 1.90 dollars per day (a derisory standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]), with a much larger share barely subsisting above that threshold and aspiring for higher and more secure living standards
[@TheWor:2018]. Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world.

At the same time, while average incomes for many countries in what was once called the "Third World" have progressively converged with those of advanced, mostly Western, economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
The widespread feelings of injustice that such unequal economic distribution arouses has historically been called the "Social Question"[@Faist:2009]. Over the past decade, inequality has become an increasingly urgent issue for policymakers as well as the general public [@Chancel:2020; @UN2020]. This urgency has been strengthened by a growing body of research demonstrating links between inequality and economic complexity, human health and wellbeing, and environmental conditions[@Hamann:2018; @Leach:2018;
@Pikett:2014; @Wilkinson:2009; @Hartmann:2017].

Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma. There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere [@Díaz:2019]. Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
[@Díaz:2019]. At the same time, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017]. Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].

Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly address prosperity, equality, and the environment [@Sachs:2015]. However, while all three of these development goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two. In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole. In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, and if it does exist, where does it occur and why does it occur there?

@@

To answer these questions, we examined the historical performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above (Table \ref{tab1}). We chose the country scale due to practical considerations
of data availability, but also on the recognition that many of the most
consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018]. We operationalized the trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g. reducing inequality and increasing prosperity) There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies are movements along
the (blue) diagonal where the achievement of two goals can be positively correlated;
while tradeoffs are represented by the (red) axis where the achievement of one
pervades the achievement of another (negative correlations). It is important to note that in this framing, the sign of the correlation changes depending on the term used, i.e. if one speaks of equality as opposed to inequality, or environmental quality as opposed to environmental footprint. In the remainder of this paper, we use inequality and environmental footprint because that is how the original data are defined and structured. Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 4in, height = 4in]{fig1_trilemma_space}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A). There could be synergies and trade-offs in achieving these three objectives (B-D). We explore trajectory typologies in the trilemma space. 53 trajectories from 1995 to 2017 are depicted. The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (E), inequality and prosperity (F), and ecological footprint and prosperity (G). Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
\label{fig:trilemma}
\end{figure*}
```

To quantify prosperity, equality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used
per-capita gross national income (GNI) in US\$ values.GNI is related to gross domestic product (GDP) and the two are close in value for most countries. A key difference is that GNI accounts for earnings from foreign investments. The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source:
<http://data.worldbank.org>). The threshold for being "prosperous" was US\$12,746 per capita per year, which is the World Bank's definition for a high-income country in 2017, the concluding year of our study period [@TheWor:2018].

To measure inequality, we used the Gini coefficient for income. The Gini coefficient is a statistical measure of dispersion and has a long history of being used to measure the level of economic inequality within a country [@Pyatt:1976]. Hypothetically, a country with a Gini value of 0 is absolutely equal, while a value of 1 represents absolute inequality. Gini coefficients of income were taken from the World Income Inequality Database v. 3.4 (source: [https://www.wider.unu.edu/data](https://www.wider.unu.edu/data)). We used 0.30 as the threshold below which countries are identified as having low inequality, similar to findings from previous historical surveys [@Piketty:2014].

We used per-capita ecological footprint to measure environmental impact. Ecological footprint is a holistic estimate of the biophysical resources used by human activity, quantified in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998]. Its unit is the "global hectare", combining measurements of carbon emissions, built-up land, fishing grounds, croplands, forests, and grazing land into an index of environmental impact. In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological "deficit". Though it is not exhaustive, the ecological footprint provides the most widely-used and data-rich approximation of overall human impact on the environment, and it accounts for the effects of trade on resource consumption by country. Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)). We used 1.68 global hectares per capita as the threshold for environmental impact; any value higher would exceed the global biocapacity (i.e., ecological deficit). In other words, it is the limit at which current generations' needs can be satisfied without compromising the ability of future generations to satisfy their needs.

@@

We used these data to chart the movement of countries across the dimensions of prosperity, equality, and environmental impact over time. We further investigated whether these movements trace out typologies of countries in the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters to fit. We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and vector-fitting to show how clusters are pulled apart in the principal component space. All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).

An analysis of the missing values of the publicly available data shows that a assessing the change over time in the trilemma spaces was only possible for 53 countries. The most incomplete dataset is the inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemmma space (Fig \ref{fig:cube_time}).

@@

No country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}F- per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but show high variation and a trend towards greater income inequality in more recent years.

While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typology. To further investigate individual trajectories, we calculated the displacement on the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane. These two angles and the total distance covered in the three-dimensional space are used to uncover typologies in country trajectories.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{trilema_ensamble}
\caption{\textbf{Countries' positions and trajectories within the trilemma space} 53 country trajectories from 1995 to 2017 are depicted. Colours and contours represent areas of similar values and enclose attractors within peaks. The insets of each graph show the direction of movement for countries during the study period.}
\label{fig:cube_ensamble}
\end{figure*}
```

There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). The inequality and ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity and ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality and prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters. Despite this clustering, countries showed distinctive profiles in terms of their direction of movement in the trilemma space (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini and ecological footprint worsened while GNI per capita grew significantly. And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories of countries in the trilemma space from 1995 to 2017 converged into three general types (Fig \ref{fig:pca}A). First, there was a group of countries "trapped" in a region of high inequality, mainly in Latin America. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$, that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). The second group showed large displacements within the trilemma space but with no definite direction, characterized by particularly high variations in Gini and per-capita GNI (Fig \ref{fig:pca}A, in yellow). The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma dimensions as summary statistics, we are able to include more countries (N=140). Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between inequality, prosperity, and environmental impact. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies, while the second (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{pca_maps}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. 3 clusters are identified when looking at the historical trajectory of countries (N=53) from 1995-2017 (A). 2 clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness. It complements the dynamic analysis with one summary statistic over time for 140 countries (B). Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
\label{fig:pca}
\end{figure*}
```

@@

Our analysis found that decreasing inequality and ecological impact pulled countries in opposite directions (Fig \ref{fig:pca}) providing strong evidence for the difficulty of achieving both goals at the same time. The historical data shows that countries typically reduce inequality and then increase environmental impact, not the other way around (Fig \ref{fig:cube_ensamble}A). Similarly, countries tend to reduce inequality first, then increase prosperity (Fig \ref{fig:cube_ensamble}B). Importantly, many middle-income countries have levels of inequality similar to high-income countries, but at lower levels of environmental impact and prosperity (Fig \ref{fig:cube_ensamble}B, C). This means that countries can achieve relatively low levels of inequality while attaining and preserving environmental integrity, but our findings show that higher prosperity often leads to higher environmental impact. On the other hand, our results also indicate that some countries struggled to escape areas in the trilemma space characterized by high levels of inequality during the period under consideration. However, it is important to note that past trajectories do not necessarily determine plausible future pathways.

The potential mechanisms of tradeoffs and synergies remain an important research frontier. Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the later grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. Since this relationship was first posited, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020]. This has made the identification of mechanisms difficult.  Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or government regulations  [@Pikett:2014]. Our analysis found no evidence for a KC-like relationship, instead finding that increasing inequality tended to precede growing prosperity.

In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase their environmental impact with growing prosperity but after reaching an inflection point of affluence, the impact of an increasingly rich society decline (Barbier 1997). As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not (Dasgupta et al. 2002; Stern 2004). In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC. Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns (Mäler 2013). But what is clear from diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.
